---
title: "Exercise for Master in personalized medicine"
author: "Rasmus F Brøndum"
date: "1/12/2021"
output: html_document
---

This document provides the basis for loading the data and training two models using the tidymodels framework. It is based a the case study from the tidymodels webpage: https://www.tidymodels.org/start/case-study/

First we load the necessary add-on packages needed for the analysis.
```{r setup, include=FALSE}
options(width = 120)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(tidymodels)
library(matrixStats)
library(vip)
```

# Load and prepare data
Load the data
```{r load_data}
## Synthetic data for analysis
df <- read_csv("cll_synth.csv")

```

You can explore the dataset by clicking on it in the menu on the right which will open a table view. The menu also shows the number of observations and variables for each dataset. You will see a number of categorical variables (e.g ECOG performance score) which have been coded as binary indicators (one-hot), and numerical values from e.g. blood measurements, which have been summarized with the minimum, maximum, standard deviation and mean value before the prediction time. The variable INFEC_after_diag is the outcome variable stating if the patient had an infection within one year of the prediction time.

You can also use the skim command to get a summary of the dataset. You may notice that some variables have no variation. These offer no value for prediction and should be filtered out prior to analysis. Furthermore, some of the numerical variables have missing values, if e.g. no blood tests were performed. One simple way to adress this is to use a median imputation, i.e. replace the missing observations with the median value for the variable in question.
```{r skim_data}
skim(df)
```

We filter out variables related to infection or not and convert the outcome to a factor. Converting to a factor with the value "1" as the first value, is a practical issue for the modelling framework which ensures that performance metrics work with default parameters.
```{r}
df <- df %>% 
  select(-c(Patient_id, infec_90_days_mean)) %>% 
  mutate(INFEC_after_diag = factor(INFEC_after_diag, levels = c("1", "0")))
```

We split the dataset into three subsets for training, validation and final test of the model. The validation set is used to tune the model hyperparameters in the training phase, while the test set is set aside for final benchmarking of the optimal model from the training.
```{r}
set.seed(42)

# Create random samples of the rows with 20% set aside for testing
df_split <- df %>% 
  initial_split(prop = 0.8)

# Extract the datasets
df_other <- training(df_split)
df_test  <- testing(df_split)

# For the non testing set we create a 75/25 train/validation split
# The original data is then split into 60/20/20 proportions for training, validation and testing
df_val <- validation_split(df_other, prop = 0.75)
```

# Elastic net Logistic Regression
Setup a model for logistic regression with elastic net. This is an extension of the traditional parametric model for logistic regression, with two tuning parameters: Penalty (lambda) which specifies how hard the model is penalized for including more features, and the mixture which sets the proportion of L1 and L2 penalty. The tuning parameteres allows the model to do variable selection, i.e. pick out the most important variables for prediction.
```{r}
lr_mod <- 
  logistic_reg(penalty = tune(), mixture = tune()) %>% 
  set_engine("glmnet")
```

Create a recipe for data pre-processing. As mentioned above we remove variables with no variation, and replace missing values with the median. For elastic net all features should be on the same scale so we normalize them to have zero mean and variance of 1.
```{r}
lr_recipe <- 
  recipe(INFEC_after_diag ~ ., data = df_other) %>% 
  step_zv(all_predictors()) %>%                # Remove variables with no variation
  step_impute_median(all_predictors()) %>%     # Impute missing values with median from training set
  step_normalize(all_predictors())             # Center and scale predictors
```

The model and recipe is combined into a workflow to be used for the parameter tuning in the training phase.
```{r}
lr_workflow <- workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(lr_recipe)
```

Setup a manual grid for hyperparameter tuning. The expand grid command, provides a sorted data.frame with all combinations of the two variables.
```{r}
lr_reg_grid <- expand_grid(penalty = 10^seq(-4, -1, length.out = 20), mixture = seq(0,1, length = 11))
```

Tune the model over the grid, i.e. train a model on the training data for each combination of the penalty and mixing parameters, and evaluate the performance on the validation data.
```{r}
lr_res <- 
  lr_workflow %>% 
  tune_grid(df_val,
            grid = lr_reg_grid,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))
```

We can print the metrics from the model training using the collect_metrics command.
```{r}
lr_res %>% 
  collect_metrics()
```

We can also plot the performance of all the models as a function of the tuning parameters, to get a quick overview.
```{r}
lr_plot <- lr_res %>% 
  collect_metrics() %>% 
  mutate(mixture = as.factor(mixture)) %>% 
  ggplot(aes(x = penalty, y = mean, color = mixture, group = mixture)) + 
  geom_point() + 
  geom_line() + 
  ylab("Area under the ROC Curve") +
  scale_x_log10(labels = scales::label_number())

lr_plot 

# A similar plot is achieved using autoplot(lr_res),
# but it is demonstrated manually for learning purposes
```

We can select the best model, i.e. combination of penalty and mixture with best AUC, by sorting on the mean AUC value, and "slicing" of the first row.
```{r}
lr_best <- 
  lr_res %>% 
  collect_metrics() %>% 
  arrange(-mean) %>% 
  slice(1)

lr_best

# Alternatively the sorting and slicing may be done using
# show_best(lr_res)
```

Finally we can plot the ROC for the best model
```{r}
lr_auc <- 
  lr_res %>% 
  collect_predictions(parameters = lr_best) %>% 
  roc_curve(INFEC_after_diag, .pred_1) %>% 
  mutate(model = "Logistic Regression")

autoplot(lr_auc)
```

# Random Forest
In this section we show an alternative model to the parametric model from the section above. In random forest we train a number of decision trees where each tree is based on a random subset (size is mtry) of the available features and split the nodes until the number of observations in is below a fixed number (min_n). Each of the trees is likely to have poor predictice performance (weak learner), but by training multiple trees and averaging across this ensemble we get a stronger learner.


We start by setting up a model for random forest, with 100 trees and tune the subset size and minimum number of observations in each node. We setup the model for classification since we are dealing with a categorical outcome.
```{r}
# This step sets up parallelization to speed up computations
cores <- parallel::detectCores()
cores

rf_mod <- 
  rand_forest(mtry = tune(), min_n = tune(), trees = 100) %>% 
  set_engine("ranger", num.threads = cores) %>% 
  set_mode("classification")
```

Create a recipe for data pre-processing similar to above, but for random forest the features do not have to be normalized.
```{r}
rf_recipe <- 
  recipe(INFEC_after_diag ~ ., data = df_other) %>% 
  step_zv(all_predictors()) %>%                      # Remove variables with no variation
  step_impute_median(all_predictors())               # Impute missing values with median from training set
```

Create a workflow
```{r}
rf_workflow <- workflow() %>% 
  add_model(rf_mod) %>% 
  add_recipe(rf_recipe)
```

Tune the model over the grid. We let the tuning framework automatically select 25 combinations of mtry and min_n.
```{r}
rf_res <- 
  rf_workflow %>% 
  tune_grid(df_val,
            grid = 25, # COuld set this manually instead
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))
```
Print the top performing models
```{r}
show_best(rf_res)
```

We can plot the performance against the tuning parameters
```{r}
autoplot(rf_res)
```

ROC for the best model
```{r}
rf_best <- 
  rf_res %>% 
  collect_metrics() %>% 
  arrange(-mean) %>% 
  slice(1)

rf_best

rf_auc <- 
  rf_res %>% 
  collect_predictions(parameters = rf_best) %>% 
  roc_curve(INFEC_after_diag, .pred_1) %>% 
  mutate(model = "Logistic Regression")

autoplot(rf_auc)
```



# Last model
This section demonstrates how to retrain the model on the combined training and validation set, using the optimal tuning parameters, and evaluating the performance on the test set.

## LR
```{r}
# Create model with best parameter
lr_final_mod <- logistic_reg(penalty = lr_best$penalty, mixture = lr_best$mixture) %>% 
  set_engine("glmnet")

# Create final workflow
lr_final_workflow <- workflow() %>% 
 add_model(lr_final_mod) %>% 
  add_recipe(lr_recipe)

# Last fit
lr_last_fit <- last_fit(lr_final_workflow, split = df_split)
lr_last_fit %>% collect_metrics()
```

We can print the most important features used for the prediction using the variable importance function. Vip depends on the type of model, so we include the type of variable importance score. For the EN model, which is a paramatric model, we get the model coefficients.
```{r}
lr_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
```

## RF
```{r}
# Create model with best parameter
# In order to extract the variable importance we need to specify it in set_engine
rf_final_mod <- 
  rand_forest(mtry = rf_best$mtry, min_n = rf_best$min_n, trees = 100) %>% 
  set_engine("ranger", num.threads = cores, importance = "impurity") %>% 
  set_mode("classification")

# Create final workflow
rf_final_workflow <- workflow() %>% 
 add_model(rf_final_mod) %>% 
  add_recipe(rf_recipe)

# Last fit
rf_last_fit <- last_fit(rf_final_workflow, split = df_split)
rf_last_fit %>% collect_metrics()
```

```{r}
rf_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
```

# Logistic Regression
Setup a model for logistic regression with elastic net. This is an extension of the traditional parametric model for logistic regression, with two tuning parameters: Penalty (lambda) which specifies how hard the model is penalized for including more features, and the mixture which sets the proportion of L1 and L2 penalty. The tuning parameteres allows the model to do variable selection, i.e. pick out the most important variables for prediction.
```{r}
lr_mod <- logistic_reg(
  mode = "classification",
  engine = "glm",
  penalty = NULL,
  mixture = NULL)
```

Create a recipe for data pre-processing. As mentioned above we remove variables with no variation, and replace missing values with the median. For elastic net all features should be on the same scale so we normalize them to have zero mean and variance of 1.
```{r}
lr_recipe <- 
  recipe(INFEC_after_diag ~ ., data = df_other) %>% 
  step_select(-c("Binet_Stage_A", "IGHV_unmut_Unmutated", "ECOG_0", "FamCLL_No",
                 "Beta2m_No", "CD38_No", "ZAP70_No", "Gender_M", "del13_No", "tri12_No", "del11_No")) %>% 
  step_zv(all_predictors()) %>%                # Remove variables with no variation
  step_impute_median(all_predictors()) %>%     # Impute missing values with median from training set
  step_normalize(all_predictors())             # Center and scale predictors
```

The model and recipe is combined into a workflow to be used in the training phase.
```{r}
lr_workflow <- workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(lr_recipe)
```

Train a model on the training data, and evaluate the performance on the validation data.
```{r}
lr_res <- 
  lr_workflow %>% 
  fit_resamples(df_val,
                control = control_resamples(
      save_pred = TRUE))
```

We can print the metrics from the model training using the collect_metrics command.
```{r}
lr_res %>% 
  collect_metrics()
```

Finally we can plot the ROC for the best model
```{r}
lr_auc <- 
  lr_res %>% 
  collect_predictions() %>% 
  roc_curve(INFEC_after_diag, .pred_1) %>% 
  mutate(model = "Logistic Regression")

autoplot(lr_auc)
```

# Last model
This section demonstrates how to retrain the model on the combined training and validation set, using the optimal tuning parameters, and evaluating the performance on the test set.

## LR
```{r}
# Create model with best parameter
lr_final_mod <- logistic_reg() %>% 
  set_engine("glm")

# Create final workflow
lr_final_workflow <- workflow() %>% 
 add_model(lr_final_mod) %>% 
  add_recipe(lr_recipe)

# Last fit
lr_last_fit <- last_fit(lr_final_workflow, split = df_split)
lr_last_fit %>% collect_metrics()
```

We can print the most important features used for the prediction using the variable importance function. Vip depends on the type of model, so we include the type of variable importance score. For the EN model, which is a paramatric model, we get the model coefficients.
```{r}
lr_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
```


```{r}
lr_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() 
```



# Gradient Boost
https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/
https://juliasilge.com/blog/xgboost-tune-volleyball/ 
```{r}
# XGBoost model specification
xgboost_model <- 
  parsnip::boost_tree(
    trees = 1000,
    min_n = tune(),
    tree_depth = tune(),
    learn_rate = tune(),
    loss_reduction = tune()
  ) %>%
    set_engine("xgboost") %>% 
  set_mode("classification")

# grid specification
xgboost_params <- 
  dials::parameters(
    min_n(),
    tree_depth(),
    learn_rate(),
    loss_reduction()
  )

xgboost_grid <- 
  dials::grid_max_entropy(
    xgboost_params, 
    size = 60
  )
```


Combine with RF recipe
```{r}
xgb_workflow <- workflow() %>% 
  add_model(xgboost_model) %>% 
  add_recipe(rf_recipe)
```

Tune
```{r}
# hyperparameter tuning
xgboost_tuned <- tune_grid(
  object = xgb_workflow,
  resamples = df_val,
  grid = xgboost_grid,
  metrics = metric_set(roc_auc),
  control = control_grid(verbose = TRUE)
)
```

