---
title: "Workshop for day 2"
author: "Inspired by Rasmus F Brøndum"
date: "23/05/2025"
output: 
  html_document:
    toc: true
    toc_float: true
---


This document provides the basis for loading the data and training some dynamic classification models using the tidymodels framework. It is based on the case study from the [tidymodels webpage](https://www.tidymodels.org/start/case-study/).

First we load the necessary add-on packages needed for the analysis.
```{r setup, include=FALSE}
options(width = 120)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(tidymodels)
library(matrixStats)
library(vip)
library(probably)
set.seed(2025)

# For possible parallelization
cores <- parallel::detectCores()
```

# Load and prepare data
Load the data
```{r load data}
# Load data from yesterday
df <- read_csv("Data_ready_for_workshop2.csv")
```

You can explore the dataset by clicking on it in the menu on the right which will open a table view. The menu also shows the number of observations and variables for each dataset. You will see a number of categorical variables (e.g treatments) which have been coded as binary indicators (one-hot), and numerical values from e.g. blood measurements, which have been summarized with the minimum, maximum, standard deviation and mean value before the prediction time. The variable mevent_nyear is the outcome variable stating if the patient had a major event in the next n years from the prediction time.

You can also use the skim command to get a summary of the dataset.
```{r skim data}
skim(df)
```

We filter out variables related to infection or not and convert the outcome to a factor. Converting to a factor with the value "1" as the first value, is a practical issue for the modelling framework which ensures that performance metrics work with default parameters.
```{r data handling}
df_cal <- df[df$split == "calibration",]
df <- df[!(df$split == "calibration"),]

# Order the splits to make them formatable in an rsplit object
df$split <- factor(df$split, levels = c("train","validation","test"))
df <- df[order(df$split),]

# Calculate the fractions of the different splits
# note the small factor to ensure the correct division
props <- c(table(df$split)[1] / sum(table(df$split)) + 1e-12,
           table(df$split)[2] / sum(table(df$split)) + 1e-12)

# Remove unnecessary columns and factorize the outcome variable
df <- df %>% 
  select(-c(id, split, date)) %>% 
  mutate(mevent_nyear = factor(mevent_nyear, levels = c("1", "0")))

df_cal <- df_cal %>% 
  select(-c(id, split, date)) %>% 
  mutate(mevent_nyear = factor(mevent_nyear, levels = c("1", "0")))
```

We split the dataset into three subsets for training, validation and final test of the model. The validation set is used to tune the model hyperparameters in the training phase, while the test set is set aside for final benchmarking of the optimal model from the training.
The calibration set is put away until later.
```{r split data}
# Split the data
df_split <- initial_validation_time_split(df, prop = props)

# Extract the datasets
df_train <- training(df_split)
df_val   <- validation_set(df_split)
df_test  <- testing(df_split)
```

The following models are going to use the same recipe, as such it is defined here.
```{r recipe}
the_recipe <- recipe(mevent_nyear ~ ., data = df_train)
the_recipe
```

# Basic Logistic Regression
Setup a basic logistic regression model. 
```{r basic LR model}
blr_mod <- logistic_reg(
  mode = "classification",
  engine = "glm",
  penalty = NULL,
  mixture = NULL)
```

The model and recipe is combined into a workflow to be used in the training phase.
```{r basic LR workflow}
blr_workflow <-
  workflow() %>% 
  add_model(blr_mod) %>% 
  add_recipe(the_recipe)
```

The model can then be evaluated using specific metrics. In this solution, we mainly consider accuracy, ROC AUC, and PR AUC, with PR being used to evaluate hyperparameters.
```{r basic LR results}
blr_res <- 
  blr_workflow %>% 
  fit_resamples(df_val,
                control = control_resamples(
                  save_pred = TRUE))
```

It is then possible to create an ROC plot.
```{r basic LR ROC plot}
blr_auc <- 
  blr_res %>% 
  collect_predictions() %>% 
  roc_curve(mevent_nyear, .pred_1) %>% 
  mutate(model = "Logistic Regression")

autoplot(blr_auc)
```

# Elastic net Logistic Regression
Setup a model for logistic regression with elastic net. This is an extension of the traditional parametric model for logistic regression, with two tuning parameters: Penalty (lambda) which specifies how hard the model is penalized for including more features, and the mixture which sets the proportion of L1 and L2 penalty. The tuning parameters allows the model to do variable selection, i.e. pick out the most important variables for prediction.
```{r LR model}
lr_mod <- logistic_reg(
  mode = "classification",
  engine = "glmnet",
  penalty = tune(),
  mixture = tune()
)
```

The model and recipe is combined into a workflow to be used for the parameter tuning in the training phase.
```{r LR workflow}
lr_workflow <-
  workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(the_recipe)
```

Setup a manual grid for hyperparameter tuning. The expand grid command, provides a sorted data.frame with all combinations of the two variables (grid search).
```{r LR grid}
lr_reg_grid <- expand_grid(penalty = 10^seq(-4, 0, length.out = 101), mixture = seq(0,1, length = 11))
```

Tune the model over the grid, i.e. train a model on the training data for each combination of the penalty and mixing parameters, and evaluate the performance on the validation data.
```{r LR tuning}
lr_res <- 
  lr_workflow %>% 
  tune_grid(df_val,
            grid = lr_reg_grid,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))
```

We can plot the performance of all the models as a function of the tuning parameters, to get a quick overview.
```{r LR hyperparameter plot}
autoplot(lr_res)
```

We can select the best model, i.e. combination of penalty and mixture with best AUC, by using the select_best command.
```{r LR best hyperparameters, warning=F}
lr_best <- lr_res %>% select_best()
```

Finally we can plot the ROC for the best model.
```{r LR ROC plot}
lr_auc <-
  lr_res %>%
  collect_predictions(parameters = lr_best) %>%
  roc_curve(mevent_nyear, .pred_1)

autoplot(lr_auc)
```

# Random Forest
In this section the Random forest model is implemented. In random forest we train a number of decision trees where each tree is based on a random subset (size is mtry) of the available features and split the nodes until the number of observations in is below a fixed number (min_n). Each of the trees is likely to have poor predictive performance (weak learner), but by training multiple trees and averaging across this ensemble we get a stronger learner.
The mentioned hyperparameters are the only ones inspected here.

We start by setting up a model for random forest, with 100 trees and tune the subset size and minimum number of observations in each node. We setup the model for classification since we are dealing with a categorical outcome.
```{r RF model}
rf_mod <- 
  rand_forest(mtry = tune(), min_n = tune(), trees = 100) %>% 
  set_engine("ranger", num.threads = cores) %>% 
  set_mode("classification")
```
Note that you could use another dataset for random forest, as it does not need normalized data.

Create a workflow.
```{r RF workflow}
rf_workflow <-
  workflow() %>% 
  add_model(rf_mod) %>% 
  add_recipe(the_recipe)
```

Tune the model over the grid. We let the tuning framework automatically select 25 combinations of mtry and min_n (random search). In practice, you would look at more than 25 combinations.
```{r RF results}
rf_res <- 
  rf_workflow %>% 
  tune_grid(df_val,
            grid = 25,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))
```

We can plot the performance against the tuning parameters
```{r RF tuning parameter plot}
autoplot(rf_res)
```

And select the best one.
```{r RF best params, warning=F}
rf_best <- rf_res %>% select_best()
```

ROC for the best model.
```{r RF ROC plot, warning = F}
rf_auc <-
  rf_res %>%
  collect_predictions(parameters = rf_best) %>%
  roc_curve(mevent_nyear, .pred_1)

autoplot(rf_auc)
```

# Extreme Gradient Boost
Next a gradient boost model is implemented, using the R-package [xgboost](https://www.r-bloggers.com/2020/05/using-xgboost-with-tidymodels/).
It is implemented using the same procedure as [Julia Silge](https://juliasilge.com/blog/xgboost-tune-volleyball/).

For the boosting model, more hyperparameters exist, compared to the random forest model.
The ones used in this solution are

- min_n: The minimum number of data points required for a split.
- loss_reduction: minimum reduction of the loss function required for a split.
- tree_depth: maximum number of splits.
- learn_rate: also known as shrinkage, changes the rate at which the model adapts.

The number of trees is set at a constant 100.
```{r GB model}
# XGBoost model specification
gb_model <- 
  boost_tree(
    trees = 100,
    min_n = tune(),
    loss_reduction = tune(),
    tree_depth = tune(),
    learn_rate = tune()
  ) %>%
  set_engine("xgboost", num.threads = cores) %>% 
  set_mode("classification")
```

Combine with the overall recipe in a workflow.
```{r GB workflow}
gb_workflow <-
  workflow() %>% 
  add_model(gb_model) %>% 
  add_recipe(the_recipe)
```

Tune using the same method as random forest.
```{r GB model tune, echo = T, results = "hide", message = F}
# hyperparameter tuning
gb_res <- 
  gb_workflow %>% 
  tune_grid(df_val,
            grid = 25,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))
```

Find the best performing hyperparameters
```{r GB best hyperparameters, warning=FALSE}
gb_best <- gb_res %>% select_best()

gb_best
```


# Neural networks
In this section, the multi-layer perceptron is implemented, also using tidymodels.
This section is loosely inspired by the [tidymodels webpage](https://www.tidymodels.org/learn/models/parsnip-nnet/).

First the model is created.
```{r Multilayer perceptron model}
mlp_mod <- mlp(
  epochs = 100,
  hidden_units = tune(),
  learn_rate = tune(),
  activation = "hardsigmoid",
  dropout = tune()) %>% 
  set_engine("brulee", num.threads = cores) %>% 
  set_mode("classification")
```

Here, in addition to the learning rate of the model, the number of hidden units and the dropout is also tuned.
Note the reason penalty is not included, as in the other models, is because penalty and dropout should not both be present.

Setup a manual grid for hyperparameter tuning. The expand grid command, provides a sorted data.frame with all combinations of the two variables.
```{r MLP grid}
mlp_reg_grid <- expand_grid(learn_rate = 10^seq(-4, -2, length.out = 3), hidden_units = c(10,25,50), dropout = c(0.1,0.5,0.9))
```

For a real analysis, you should consider using a finer grid, a wider range, more hidden layers, and more tuning parameters.
For inspiration on which hyperparameters to choose, see the [supplementary material](https://ascopubs.org/action/downloadSupplement?doi=10.1200%2FCCI.22.00054&file=DS_CCI.22.00054.pdf) for the paper this course is based on.

Then the workflow has to be defined, using the same predictors as the other models.
```{r MLP workflow}
mlp_workflow <- 
  the_recipe %>% 
  workflow(mlp_mod)
```

The model has three hyperparameters that need estimating, this is done over the above defined grid.
```{r MLP hyperparameters}
mlp_res <-
  mlp_workflow %>%
  tune_grid(df_val,
            grid = mlp_reg_grid,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(roc_auc))
```

Plot the different values for the hyperparameters.
```{r MLP hyperparameter plot}
mlp_res %>% autoplot()
```

As ROC is used to evaluate the hyperparameters, the ROC plot is shown.
```{r MLP ROC plot, warning = F}
mlp_best <- mlp_res %>% select_best()

mlp_auc <-
  mlp_res %>%
  collect_predictions(parameters = mlp_best) %>%
  roc_curve(mevent_nyear, .pred_1)

autoplot(mlp_auc)
```


# Final models
This section demonstrates how to retrain the model on the combined training and validation set, using the optimal tuning parameters, and evaluating the performance on the test set.

## Basic logistic regression
```{r basic LR final model, warning = F}
# Create model with best parameter
blr_final_mod <- logistic_reg() %>% 
  set_engine("glm")

# Create final workflow
blr_final_workflow <- workflow() %>% 
  add_model(blr_final_mod) %>% 
  add_recipe(the_recipe)

# Last fit
blr_last_fit <- last_fit(blr_final_workflow, split = df_split,
                         metrics = metric_set(precision, roc_auc, pr_auc, brier_class),
                         add_validation_set = F)
blr_last_fit %>% collect_metrics()
```
We can print the most important features used for the prediction using the variable importance function. Vip depends on the type of model, so we include the type of variable importance score. For the logistic model, which is a parametric model, we can use the model coefficients (or something based on them).
```{r basic LR predictor importance}
blr_vars <- blr_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
blr_vars
```

## Logistic regression with elastic net
```{r LR final model}
# Create model with best parameter
lr_final_mod <- logistic_reg(penalty = lr_best$penalty, mixture = lr_best$mixture) %>% 
  set_engine("glmnet")

# Create final workflow
lr_final_workflow <- workflow() %>% 
  add_model(lr_final_mod) %>% 
  add_recipe(the_recipe)

# Last fit
lr_last_fit <- last_fit(lr_final_workflow, split = df_split,
                        metrics = metric_set(precision, roc_auc, pr_auc, brier_class),
                        add_validation_set = T)
lr_last_fit %>% collect_metrics()
```

The elastic net does not remove the fact that it is a parametric model, so similar methods can be used.
```{r LR predictor importance}
LRvars <- lr_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
LRvars
```

## Random forest
Next consider the random forest model, where rf_best contains the best hyperparameters.
```{r RF final mod}
# In order to extract the variable importance we need to specify it in set_engine
rf_final_mod <- 
  rand_forest(mtry = rf_best$mtry, min_n = rf_best$min_n, trees = 100) %>% 
  set_engine("ranger", num.threads = 1, importance = "impurity") %>% 
  set_mode("classification")

# Create final workflow
rf_final_workflow <- workflow() %>% 
  add_model(rf_final_mod) %>% 
  add_recipe(the_recipe)

# Last fit
rf_last_fit <- last_fit(rf_final_workflow, split = df_split,
                        metrics = metric_set(precision, roc_auc, pr_auc, brier_class),
                        add_validation_set = T)
rf_last_fit %>% collect_metrics()
```

As random forests are not a parametric model, the important predictors are instead based on impurity.
```{r RF predictor importance}
rf_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
```

## Gradient Boost
Next, consider the gradient boost model.
```{r GB final mod}
gb_final_mod <- boost_tree(
  trees = 250,
  min_n = gb_best$min_n,
  tree_depth = gb_best$tree_depth,
  learn_rate = gb_best$learn_rate,
  loss_reduction = gb_best$loss_reduction
) %>%
  set_engine("xgboost", importance = "impurity", num.threads = cores) %>% 
  set_mode("classification")

# Create final workflow
gb_final_workflow <- workflow() %>% 
  add_model(gb_final_mod) %>% 
  add_recipe(the_recipe)

# Last fit
gb_last_fit <- last_fit(gb_final_workflow, split = df_split,
                        metrics = metric_set(precision, roc_auc, pr_auc, brier_class),
                        add_validation_set = T)
gb_last_fit %>% collect_metrics()
```

The vip is now based on the amount of gain each coefficient contributed with (vaguely).
```{r GB predictor importance}
gb_last_fit %>% 
  pluck(".workflow", 1) %>%   
  extract_fit_parsnip() %>% 
  vip(num_features = 20, include_type = T)
```

## Multi-layer perceptron
Finally, consider the multi-layer perceptron
```{r MLP final mod}
mlp_final_mod <- mlp(
  epochs = 100,
  hidden_units = mlp_best$hidden_units,
  dropout = mlp_best$dropout,
  learn_rate = mlp_best$learn_rate,
  activation = "hardsigmoid") %>% 
  set_engine("brulee", num.threads = cores) %>% 
  set_mode("classification")

# Create final workflow
mlp_final_workflow <- workflow() %>% 
  add_model(mlp_final_mod) %>% 
  add_recipe(the_recipe)

# Last fit
mlp_last_fit <- last_fit(mlp_final_workflow, split = df_split,
                         metrics = metric_set(precision, roc_auc, pr_auc, brier_class),
                         add_validation_set = T)
mlp_last_fit %>% collect_metrics()
```
For this type of model, variable importance is not easily calculated, consider SHAP values if need be.

# Results
Summarise all of the different model results in a single data frame.
```{r Results}
metric_vals <- cbind(
  blr_last_fit %>% collect_metrics() %>% select(c(.metric, .estimate)) %>% rename(basic_LR = .estimate),
  lr_last_fit  %>% collect_metrics() %>% select(.estimate) %>% rename(LR = .estimate),
  rf_last_fit  %>% collect_metrics() %>% select(.estimate) %>% rename(RF = .estimate),
  gb_last_fit  %>% collect_metrics() %>% select(.estimate) %>% rename(boost = .estimate),
  mlp_last_fit %>% collect_metrics() %>% select(.estimate) %>% rename(MLP = .estimate)
)
metric_vals
```

# Model calibration
Now that the models have been created and compared, it is important to calibrate them, such that the given risk scores are meaningful.
In this case, for brevity, only the random forest model is calibrated
```{r gather the predictions for the test data}
# random forest predictions on test data
rf_pred <- rf_last_fit %>% 
  collect_predictions() %>% 
  select(rf = .pred_1) %>% 
  cbind(mevent_nyear = df_test$mevent_nyear)
# and the corresponding predictions for the calibration data set
rf_cal_pred <- rf_last_fit %>% 
  extract_workflow() %>% 
  predict(new_data = df_cal, type = "prob") %>% 
  select(rf = .pred_1) %>% 
  cbind(mevent_nyear = df_cal$mevent_nyear)
```

Next the calibration plot is created.
```{r pre-calibration plot}
# Creates a calibration plot using the "probably" package
rf_pred %>% cal_plot_breaks(mevent_nyear,  rf)
```

Isotonic regression typically overfits, and we are dealing with a small data set.
As such only Platt scaling is considered (even though this also has some problems with imbalanced data sets).
```{r Platt calibration}
# the calibration model
rf_platt <-
  workflow() %>% 
  add_model(blr_mod) %>% 
  add_recipe(recipe(mevent_nyear ~ rf, data = rf_pred)) %>%
  fit(data = rf_pred)
# the resulting calibrated probabilities
rf_platt_test <- rf_platt %>%
  predict(new_data = rf_pred, type = "prob") %>% 
  select(.pred_1) %>% 
  unlist()
# the calibration is validated on the calibration data set
rf_platt_cal <- rf_platt %>% 
  predict(new_data = rf_cal_pred, type = "prob") %>% 
  select(.pred_1) %>% 
  unlist()
```

Compare the calibrated values, to the uncalibrated ones.
```{r Brier score for calibrated models compared}
# Compare them to the Brier score for the uncalibrated values
final_cal <- cbind(Uncalibrated = metric_vals[4,4],
                   Calibrated = brier_class_vec(truth = df_test$mevent_nyear, estimate = rf_platt_test))

final_cal
```

Next consider the calibration plots.
```{r post-calibration plot}
# Creates a calibration plot using the "probably" package
cbind(rf = rf_platt_test, mevent_nyear = (df_test$mevent_nyear==1)) %>%
  as.data.frame() %>%
  cal_plot_breaks(mevent_nyear,  rf)
```

Then check how it looks for the calibration set.
```{r calibration validation}
# Creates a calibration plot using the "probably" package
cbind(rf = rf_platt_cal, mevent_nyear = (df_cal$mevent_nyear==1)) %>%
  as.data.frame() %>%
  cal_plot_breaks(mevent_nyear,  rf)
```
# Save a model for day 3
```{r saving a model}
saveRDS(rf_last_fit, file = "trained_model.RDS")
```


