---
title: "Workshop 0: Data Sources and Data Acquisition Exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Data Sources and Data Acquisition Exercise

Tasks:

1) **Create Metadata**: Write metadata files for the following datasets:

- `baseline_data.csv`
- `blood_data.csv`
- `diag_data.csv`

Optional: Create metadata for all other .csv files located in the /Data directory.

2) **Translate SKS to ICD-10**: In the `dict_data.csv` dataset, translate the SKS codes into ICD-10 codes. Create a script that checks the SKS code and adds the corresponding ICD-10 code as a new column.

Hint: The ICD-10 codes can be found on this database site: https://icd.who.int/browse10/2019/en.

Optional: Make the same script for the `diag_data.csv`.

## 1) Create Metadata

```{r}
baseline_data <- read.csv("data/baseline_data.csv")
blood_data <- read.csv("data/blood_data.csv")
diag_data <- read.csv("data/diag_data.csv")
dict_data <- read.csv("data/dict_data.csv")
events <- read.csv("data/events.csv")
```


```{r}
head(baseline_data)
```

```{r}
cat(paste(rep("=", 100), collapse = ""), "\n")
cat("Metadata DataFrame for: baseline_data\n")
cat(paste(rep("=", 100), collapse = ""), "\n")

baseline_metadata_df <- data.frame(
  `Column Name` = names(baseline_data),
  `Data Type` = sapply(baseline_data, class),
  row.names = NULL
)

baseline_descriptions <- c(
  "id" = "Unique identifier for each patient",
  "sex" = "Biological sex of the patient",
  "d.birth" = "Patient's date of birth",
  "smoking" = "Smoking status of the patient (current, former, never)"
)

baseline_metadata_df$Description <- baseline_descriptions[baseline_metadata_df$Column.Name]
print(baseline_metadata_df)
cat("\n")


# Metadata for dict_data
cat(paste(rep("=", 100), collapse = ""), "\n")
cat("Metadata DataFrame for: dict_data\n")
cat(paste(rep("=", 100), collapse = ""), "\n")

dict_metadata_df <- data.frame(
  `Column Name` = names(dict_data),
  `Data Type` = sapply(dict_data, class),
  row.names = NULL
)

dict_descriptions <- c(
  "X" = "Index for the diagnosis",
  "diag" = "Name of the diagnosis",
  "code" = "Diagnostic code (SKS)"
)

dict_metadata_df$Description <- dict_descriptions[dict_metadata_df$Column.Name]
print(dict_metadata_df)
cat("\n")


# Metadata for events
cat(paste(rep("=", 100), collapse = ""), "\n")
cat("Metadata DataFrame for: events\n")
cat(paste(rep("=", 100), collapse = ""), "\n")

events_metadata_df <- data.frame(
  `Column Name` = names(events),
  `Data Type` = sapply(events, class),
  row.names = NULL
)

events_descriptions <- c(
  "X" = "Index for the diagnosis",
  "id" = "Unique identifier for each patient",
  "date" = "Date the event occurred",
  "event" = "Description of the medical event, e.g Malignancy",
  "type" = "Classification of the event type (minor, major, death)"
)

events_metadata_df$Description <- events_descriptions[events_metadata_df$Column.Name]
print(events_metadata_df)
```

## 2) Translate SKS to ICD-10

```{r preview-dict_data}
dict_data
```

```{r translate-sks}
dict_data <- dict_data %>%
  mutate(
    `ICD-10_code` = case_when(
      code == "DE10"   ~ "E10",
      code == "BZFC8A" ~ "E78.5",
      code == "DE782"  ~ "I10",
      TRUE             ~ NA_character_
    )
  )

dict_data
```

## 3) Find Danish Registries
